Bayesian exponential random graph modeling of KWL Burial network

### 2020-11-12 analysis/paper/paper.Rmd
add references and edit Intro

<!--burial study- network analysis-->
Burials analysis by archaeologists is an approach to understand past social structure through the study of the physical traces of mortuary practices. The material cultures and biological records of burial behaviors can represent the social status or identity of the deceased [@Binford1971; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. For example, mortuary practices, including burial forms, grave goods, and ritual behaviors, that structure the material configuration of burials can represent social relations between members of a community, allowing the inference of social structures [@Byrd1995; @Seikel2011]. The recent development of network science provides many new tools to visualize and analyze relationships among ancient burials. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation, based on our specification in terms of network theories [@Brandes2013]. That is, a network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data.

<!-- intro to networks -->
A network is generally visualized as a graph consisting of a set of socially-relevant nodes, or actors, connected by edges representing one or more relations, such as friendship, collaborations, information flow, trade ties, or any other forms of connection of interest [@Marin2011; @Wasserman1994]. In archaeology, actors can be people, groups, objects, places, or events, with ties built on similarity, proximity, and co-presence of material culture based on theories of relationships [@Mills2017; @Brughmans2018]. For example, past trade can be conceptualized as a network of individual entities connected by shared similarity, the flow of goods, to represent their interactions [@Collar2015, pp. 4]. Similar concepts can be applied to a wide range of archaeological data with relational assumptions, such as burial contexts. Burial goods, especially high value goods, can reflect social practices in broader cultural contexts to represent personal wealth or social status from which we can infer social differentiation or complexity [@Gamble2002; @Janes2013]. Burials with the same prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This enables the exploration of the structure of the past social organization through the identification of the relationships among burials.

<!-- two types of network analysis -->
Network analysis has been increasingly applied by archaeologists in recent years to deal with past interactions and explore the underlying mechanisms. There are two common approaches, one is directly examining network properties, such as node position and overall structures of networks built on archaeological data, to find meaningful patterns [@Mills2017]. Another approach is using models and simulations to represent and test dependence of relationships [@Brughmans2018]. Node position is important to identify the role of a node in a network, such as its centrality, representing the individual influence or social prominence in a group, while network structure is helpful to generalize relationship patterns. Recent applications include the investigation of long term inter-site relationships from the Epipalaeolithic to the early Neolithic in the Near East according to trade items [@Coward2013]. Similarly, @Mizoguchi2013 analyzed hierarchical communication systems in the Kofun period in Japan through the study of prestige goods. Regarding burials, @Sosna examined spatial pattern of burials from the Early Bronze Age in Rebesovice with two hypothesized networks constructed according to cultural and chronological differences between burials. They found that the observed concentric pattern does not match the network results projected into geographic space, but contemporary subsections that indicates different corporate groups. For the second approach, researchers typically statistically test hypotheses about the relationships between actors to examine the structure of the network and infer major factors that affect the formation of the structure [@Freeman2004; @Salvini2010]. Examples include accessing hunter-gatherer exchange networks structure across the Kuril Islands using bootstrap simulation based on ceramic composition [@Erik2015]. Another example is to simulate several modes of food exchange for Ancestral Pueblos and compare the results with the social networks build on the aggregation of households in the American Southwest at a regional scale [@Crabtree2015]. This paper will use a novel Bayesian modeling approach to investigate the relationship between burials in northeastern Taiwan to explore the indirect impact of foreign contacts on social organisation.

<!-- archaeological background -->
Archaeological sites in northeastern Taiwan show evidence of use of imported prestige goods in burials and some accumulation patterns in residential areas around the same time as European contact in the 17th century [@Chen2007]. This might indicate the pursuit of prestige or wealth in those Indigenous societies. Prestige goods include imported ornaments and porcelains that entered northeastern Taiwan via regional exchange networks that involved the Dutch, Spanish, Han Chinese traders, and local Indigenous groups. The use of prestige goods may have began in local Indigenous communities before a European presence became established, but was amplified during the European colonial period [@Borao2009; @LiandWu2006; @LWandBM2020ornament]. For example, an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status by their community [@LiandWu2006]. A general model to summarize this situation in northeastern Taiwan might be that the influence of a colonial power, combined with high local values of imported goods, might lead to increased social inequality due to competition among individuals, and unequal access to trade networks <!-- citation -->.

<!-- archaeological specific case study -->
In this study, we use a pericolonial model to explore the changes in social inequality at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century <!-- citation -->. Our purpose is to examine the network structure of burials before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the of network our hypothesized model? (2) did European colonial activities in 17th century Taiwan result in the emergence of social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of burials in understanding the indirect effects of a colonial presence on Indigenous groups.

>>>>

<!--burial study- network analysis-->
Burials analysis by archaeologists is an approach to understand past social structure through the study of the physical traces of mortuary practices. The material cultures and biological records of burial behaviors can represent the social status or identity of the deceased [@Binford1971; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. For example, mortuary practices, including burial forms, grave goods, and ritual behaviors, that structure the material configuration of burials can represent social relations between members of a community, allowing the inference of social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theories, the recent development of social network analysis provides many new tools to visualize and analyze relationships for archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation, based on our specification in terms of network theories [@Brandes2013; @Collar2015]. That is, a network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data.

<!-- intro to networks -->
A social network is generally visualized as a graph consisting of a set of socially-relevant nodes/actors, connected by edges/ties representing one or more relations, such as friendship, collaborations, information flow, trade ties, or any other forms of connection of interest [@Marin2011; @Wasserman1994]. The ties can be classified into four major types, including similarities, social relations, interactions, and flows [@Borgatti2009]. In archaeology, actors can be people, groups, objects, places, or events, with ties built on similarity, proximity, or co-presence of material culture to create patterns reflecting influence, geographical distance, or affiliations in social groups [@Mills2017; @Brughmans2018; @Peeples]. For example, past trade can be conceptualized as a network of individual entities connected by shared similarity, the flow of goods, to represent their interactions [@Collar2015, pp. 4]. Similar concepts can be applied to a wide range of archaeological data with relational assumptions, such as burial contexts. Burial goods, especially high value goods, can reflect social practices in broader cultural contexts to represent personal wealth or social status from which we can infer social differentiation or complexity [@Gamble2002; @Janes2013]. Burials with the same prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This enables the exploration of the structure of the past social organization through the identification of the relationships among burials.

<!-- two types of network analysis -->
Network analysis has been increasingly applied by archaeologists in recent years to deal with past interactions and explore the underlying mechanisms. There are two common approaches to characterize network properties at two distinct scales: node/edge level and graph level [@Peeples]. Node level focuses on the role of node in a network, such as centrality, representing the individual influence or social prominence in a group, while graph level assesses the whole network attributes, such as density, clustering in a network, to generalize relationship patterns [@Mills2017; @Peeples]. By quantifying those network properties, archaeologists can answer a wide range of research questions. Examples includes exploring the political centralization in the Kofun period in Japan through the hierarchical communication network constructed by prestige goods [@Mizoguchi2013], the investigation of long term inter-site relationships from the Epipalaeolithic to the early Neolithic in the Near East according to trade items [@Coward2013]. Regarding burials, @Sosna2013 examined spatial pattern of burials from the Early Bronze Age in Rebesovice with two hypothesized networks constructed according to cultural and chronological similarity between burials. Recently, complex network modeling evaluates networks at both node and graph level through simulations of particular processes and statistically testing the formation of network properties [@Brughmans2013; @Brughmans2018; @Freeman2004; @Salvini2010]. Such application includes simulations and testing food exchange modes for Ancestral Pueblos on the aggregation of households in the American Southwest simulation [@Crabtree2015], or exploring the diffusion of fired bricks across Hellenistic Europe by comparing similarity networks of sites with random networks. Other examples show the assessment of hunter-gatherer exchange networks structure across the Kuril Islands using bootstrap simulation based on ceramic composition [@Erik2015]. This paper will use a novel Bayesian modeling approach to investigate the formation of relationship between burials in northeastern Taiwan to explore the indirect impact of foreign contacts on social organization.

<!-- archaeological background -->
Archaeological sites in northeastern Taiwan show evidence of use of imported prestige goods in burials and some accumulation patterns in residential areas around the same time as European contact in the 17th century [@Chen2007]. This might indicate the pursuit of prestige or wealth in those Indigenous societies. Prestige goods include imported ornaments and porcelains that entered northeastern Taiwan via regional exchange networks that involved the Dutch, Spanish, Han Chinese traders, and local Indigenous groups. The use of prestige goods may have began in local Indigenous communities before a European presence became established, but was amplified during the European colonial period [@Borao2009; @LiandWu2006; @LWandBM2020ornament]. For example, an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status by their community [@LiandWu2006]. A general model to summarize this situation in northeastern Taiwan might be that the influence of a colonial power, combined with high local values of imported goods, might lead to increased social inequality due to competition among individuals, and unequal access to trade networks [@Brumfiel1994; @Clark1994].

<!-- archaeological specific case study -->
In this study, we use a pericolonial model to explore the changes in social inequality at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. Our purpose is to examine the network structure of burials before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the of network our hypothesized model? (2) did European colonial activities in 17th century Taiwan result in the emergence of social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of burials in understanding the indirect effects of a colonial presence on Indigenous groups.
